---
title: "Powerlifting: An Exploration on the Data"
author: "Zach Eanes & Dalton Rogers"
output: powerpoint_presentation
date: "2024-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

ipf <- read_csv("openipf-2024-04-13.csv")
# View(ipf)

ipf <- janitor::clean_names(ipf)
```

# What is "Powerlifting"?
As an easy description, powerlifting is the open sport of competing to see who can lift
the most total weight across three primary lifts. These lifts being Squat, Bench, and Deadlift 
(referred to as SBD throughout this project). 

# The general structure of many of these competitions includes many little factors:
1) Everyone is split into weight classes, gender, and age classes. This is to ensure everyone is 
   competing fairly and no one has a clear advantage.
2) Every lifter has 3 attempts for each lifting, leaving 9 total lifts across the competition. 
3) Drug free! No steroids or substances that may boost someone's performance, and these 
   competitors are tested to ensure they are drug free.
4) The concept of good and failed lifts: There's 3 judges for a lift, all viewing a different 
   perspective of the lifter. In order for a lift to be good, 2/3 judges must deem it a good lift.
   Some common failures include not getting enough depth squatting, not controlling the drop after
   a deadlift, or of course literally not getting the weight up.

# Linear Progression of Weight Classes? (Section 1)
This one is simple, do heavier weights classes generally lift higher amounts than lower? We can
observe the best successful lift for each SBD and find an average for each weight class group
and then we can compare the classes to see if there's a linear progression.

# Weight Classes for IPF
IPF lists their male weight classes as:
59.0 kg Class:   up to 59.0 kg
66.0 kg Class:   from 59.01 kg up to 66.0 kg
74.0 kg Class:   from 66.01 kg up to 74.0 kg
83.0 kg Class:   from 74.01 kg up to 83.0 kg
93.0 kg Class:   from 83.01 kg up to 93.0 kg
105.0 kg Class:  from 93.01 kg up to 105.0 kg
120.0 kg Class:  from 105.01 kg up to 120.0 kg
120.0+ kg Class: from 120.01 kg up to unlimited

IPF lists their female weight classes as:
47.0 kg Class:  up to 47.0 kg
52.0 kg Class:  from 47.01 kg up to 52.0 kg
57.0 kg Class:  from 52.01 kg up to 57.0 kg
63.0 kg Class:  from 57.01 kg up to 63.0 kg
69.0 kg Class:  from 63.01 kg up to 69.0 kg
76.0 kg Class:  from 69.01 kg up to 76.0 kg
84.0 kg Class:  from 76.01 kg up to 84.0 kg
84.0+ kg Class: from 84.01 kg up to unlimited

# Functions Used
```{r avg_success_functions}

# This function gets the average of the best lift for each factor and arranges it
# by the attribute(s) passed in
#
# Args: class - the class we want to get the average weights for
#       grouping_by - the attribute(s) to group and arrange the data by
avg_weights <- function(class, grouping_by) {
    class |> 
    filter(!is.na(best3squat_kg) & !is.na(best3bench_kg) & !is.na(best3deadlift_kg) & !is.na({{grouping_by}})) |> 
        group_by({{grouping_by}}) |> # group by their respective classes
        summarize(
            avg_squat = sum(best3squat_kg) / n(), # get avg squat
            avg_bench = sum(best3bench_kg) / n(), # get avg bench
            avg_dead = sum(best3deadlift_kg) / n()) |> # get avg deadlift 
        arrange({{grouping_by}}) # arrange to make it better
}

# This function takes our data and makes the data itself tidier and simple to use for
# both data exploration and graphing.
#
# Args: class - the class we want to make tidier 
make_avgs_tidyer <- function(class) {
    # pivot to long format to graph easier
    class |> pivot_longer(cols = c(avg_squat, avg_bench, avg_dead),
                          names_to = "lift", values_to = "avg_weight")
}
```

# Male Weight Classes
``` {r} 
# get those in proper weight classes, and make weight_class_kg a factor for organization
male_factors = c(59, 66, 74, 83, 93, 105, 120, "120+")
male_classes <- ipf |> 
    filter(weight_class_kg %in% c(59, 66, 74, 83, 93, 105, 120, "120+") & sex == "M") |>
    mutate(weight_class_kg = factor(weight_class_kg, levels = male_factors))

# get information we need for our data and make it usable
weights_male <- avg_weights(male_classes, weight_class_kg) |> make_avgs_tidyer()
weights_male
# group by the weight class and put sbd all next to each other in group
weights_male |> ggplot(aes(x = weight_class_kg, y = avg_weight, fill = lift)) + 
    geom_bar(stat = "identity", position = "dodge") +
    labs(
        title = "Comparing Male Average Weights for SBD per Weight Class",
        x = "Weight Class (kg)",
        y = "Average Weight (kg)"
    ) + scale_fill_discrete(name = "What Lift?", labels = c("Bench", "Deadlift", "Squat"))

```

# Female Weight Classes
``` {r} 
# get those in proper weight classes, and make weight_class_kg a factor for organization
female_factors = c(47, 52, 57, 63, 69, 76, 84, "84+")
female_classes <- ipf |>
    filter(weight_class_kg %in% c(47, 52, 57, 63, 69, 76, 84, "84+") & sex == "F") |>
    mutate(weight_class_kg = factor(weight_class_kg, levels = female_factors))
          
# get information we need for our data and make it usable
weights_female <- avg_weights(female_classes, weight_class_kg) |> make_avgs_tidyer() 

# group by the weight class and put sbd all next to each other in group
weights_female |> ggplot(aes(x = weight_class_kg, y = avg_weight, fill = lift)) + 
    geom_bar(stat = "identity", position = "dodge") +
    labs(
        title = "Comparing Female Average Weights for SBD per Weight Class",
        x = "Weight Class (kg)",
        y = "Average Weight (kg)"
    ) + scale_fill_discrete(name = "What Lift?", labels = c("Bench", "Deadlift", "Squat"))
```

# Failures...
In order to get this information together, we had many failures and attempts at using the 
data is a less tidy manner. The biggest change was the inclusion of the pivot_longer in order
to allow us to create more rows and make the graphing easier. It allows us to group and place
each lift inside of the group, as well as creating a cleaner chart in general. Beforehand,
we broke some of the basic policies of tidy data, including storing too much information
inside of a cell and "double dipping". We had numbers in the columns for the lift itself, and
in the end out newest form is tidiest and surely best for our purposes. Below is some of the 
attempts we had beforehand, which ended as a failure in all honesty.

# R Code for Failure
```{r avg_failure}
# recreate our data set for demonstration purposes
male_classes <- ipf |> 
    filter(weight_class_kg %in% c(59, 66, 74, 83, 93, 105, 120, "120+") & sex == "M") |>
    mutate(weight_class_kg = factor(weight_class_kg, 
                                       levels = male_factors))
# this was the dataset before
avg_weights(male_classes) 

# this was out plot before the pivot_longer, as we just had each lift in thier own graph
avg_weights(male_classes, weight_class_kg) |> ggplot(aes(x = weight_class_kg, y = avg_squat, fill = avg_squat)) + 
    geom_bar(stat = "identity") 

```

# Conclusion for Section 1
Overall, we were able to find that there is a general linear progression of weight classes
and the amount of weight lifted. This is a trend we expected, as the heavier weight classes
generally have more muscle mass and are able to lift more weight.

Interestingly, we found that the womens 76kg class completely beat out the class above them, 
the 84kg. Along with this, towards the higher classes, there's a slight taper on the 
deadlifts, and our guess is because of grip strength limitations.

# Percentage of Failed Lifts Across all Nine Attempts (Section 2)
In this section, we wanted to observe if for certain lifts, competitors are more likely to
fail the lift. In the database, if a life begins with a "-", this means it was a failure. 
Using this, we may be able to get the percentage of good lifts to failed lifts for each lift, 
all 3 for SBD.

# Setup for Graphing
```{r}
# first we remove bad entries, where any of their lifts are na. sum up only data we want
# to observe
good_lifts <- ipf |> 
    # filter our bad entries
    filter(!is.na(squat1kg) & !is.na(squat2kg) & !is.na(squat3kg) &
           !is.na(bench1kg) & !is.na(bench2kg) & !is.na(bench3kg) &
           !is.na(deadlift1kg) & !is.na(deadlift2kg) & !is.na(deadlift3kg)) |> 
    # only get what we need
    summarize(name, squat1kg, squat2kg, squat3kg, bench1kg, bench2kg, bench3kg,
              deadlift1kg, deadlift2kg, deadlift3kg)
    
# pivot longer to get the data as column lift with a value of the column name and weight 
# as a value in another
good_lifts <- good_lifts |> 
    # split the dataset into lift and weight, to observe entries as group easier
    pivot_longer(cols = c(squat1kg, squat2kg, squat3kg, bench1kg, bench2kg, bench3kg, 
                          deadlift1kg, deadlift2kg, deadlift3kg),
                 names_to = "lift", values_to = "weight")
```

# Getting the Percentage of Failed Lifts
```{r} 
# this function gets the percentage of failed lifts for each lift. because the dataset 
# lists the failed lifts beginning with a -, we can use regex to explore failure 
# to success
#
# Args: class - the data we want to get the percentage of failed lifts for
percentage_failed <- function(class) {
    class |> 
        # group by lifts to see the success/fail rate of each
        group_by(lift) |>
        # get the percentage of failed lifts for each lift
        summarize( # get the lift and their respective failure/success rate
            failed = sum(str_detect(weight, "^-.*")) / n() * 100,
            num_failed = sum(str_detect(weight, "^-.*")),
            success = 100 - failed,
            num_success = n() - num_failed,
            total = failed + success
        )
}

# get the percentage of failed lifts for each lift
rates <- percentage_failed(good_lifts)
```

# Graphing the Data
```{r}
# longer data in order to graph easier :)
lift_data_long <- rates |> 
    pivot_longer(cols = c(failed, success), names_to = "outcome", values_to = "percentage")

# Plot
ggplot(lift_data_long, aes(x = lift, y = percentage, fill = outcome)) +
    # bar chart of the percentage of failed and successful lifts for each lift
    geom_bar(stat = "identity") + 
    # add text to the chart showing the percentage itself 
    geom_text(aes(label = paste0(round(percentage, 2), "%")), 
              position = position_stack(vjust = 0.5), 
              color = "white") +
    labs(
        title = "Percentage of Failed and Successful Lifts for Each Lift",
        x = "Lift",
        y = "Percentage (%)",
        fill = "Outcome"
    ) +
    # colors for the percentages themselves 
    scale_fill_manual(values = c("success" = "darkgreen", "failed" = "red"), 
                      name = "Outcome of Lift", labels = c("Failure", "Success")) +
    # make the x axis text easier to read
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Conclusion for Section 2
Looking at the chart itself, there's some shocking results. First, we never expected the last
bench attempt to have the highest failure rate. This is because the bench press is generally
attempted safely, as competitors still have the deadlift after to try and get more. Our guess was
that the last deadlift would have the highest failure rate, for many reasons:
1) It's the last lift, so competitors will be more fatigued from the rest of the event. 
2) Since it's the last, many competitors go "all-out" to attempt to place on the podium.
3) In generality, deadlift is the hardest lift to get up, because grip strength is a large 
   limiting factor for the strongest competitors.
   
Lastly, we were able to conclude that there is some trend to where the first lift is the most
successful, and the last lift is the least successful. This is because the first lift is generally
the lightest, and the last lift is generally the heaviest. This is a trend we expected, but it's
nice to see it confirmed in the data.


# Is Powerlifting a growing sport? (Section 3)

In this section, we simply want to see if the data supports the idea that powerlifting is a
growing sport. It's been said to be a growing sport, and many say that the explosion of social 
media especially backs this up. To do this, we can observe the total amount of competitors in 
each year, and see if there's a trend of growth. 

# Questions to Answer
There's also some other interesting ideas we'd like to see:
1) Did the pandemic of COVID-19 have an effect on the amount of competitors? If so, what years 
   were most affected by it?
2) Which years had the largest percent growth from the previous?
3) Which years had the largest percent decrease from the previous?
4) What is the average growth rate of competitors per year?

By being able to look at the data to support this idea, especially for the largest powerlifting 
federation, IPF, we can see if the sport is truly growing or not. 

# Calculations for our Questions
```{r}
# using the date column, get the year of each entry
years_ipf <- ipf |> mutate(year = year(date)) |> group_by(year) |> arrange(year)

# get the total amount of competitors for each year 
total_competitors <- years_ipf |> summarize(total_competitors = n())

# get the percent growth from the previous year
total_competitors <- total_competitors |> 
    mutate(percent_growth = (total_competitors - lag(total_competitors)) / lag(total_competitors) * 100)

# get the average growth rate of competitors per year
avg_growth_rate <- total_competitors |>
    summarize(avg_growth_rate = mean(percent_growth, na.rm = TRUE))

avg_growth_rate
```

# Graphing the Data
```{r}
# plot the total competitors per year
total_competitors |> ggplot(aes(x = year, y = total_competitors)) +
    # line to show the true trend of competitors
    geom_line() + 
    # bar chart to show individual years clearer
    geom_bar(stat = "identity", alpha = 0.5, fill = "purple") +
    labs(
        title = "Total Amount of Competitors per Year",
        x = "Year",
        y = "Total Competitors"
    )
```

# Percentage Growth of Each Year
```{r}
# plot the percentage growth of each year
total_competitors |> ggplot(aes(x = year, y = percent_growth)) +
    # line to show the true trend of growth
    geom_line() + 
    # bar chart to show individual years clearer
    geom_bar(stat = "identity", alpha = 0.5, fill = "blue") +
    labs(
        title = "Percentage Growth of Competitors per Year",
        x = "Year",
        y = "Percentage Growth (%)"
    )
```

# Conclusion for Section 3
After our analysis, we we able to conclude that in general powerlifting is a growing sport. 
We can see this is the general trend of it growing upwards, as well we found the average growth
rate for each year to be roughly 33.5%. Overall, there is a very positive trend for powerlifting
and it is growing more than ever. 

As we guessed, COVID had a massive impact on the growth and affect of IPF competitor counts. From 
our data, the year 2020 and 2021 both had large drop-offs from the previous 2019, where the sport
seemed to have peaked. Interestingly, the largest percentage growth comes in the earlier years as 
well, such as 1979 and 1980. After a short google search, we were able to find that the IPF 
world championships were held in Dayton, Ohio. Since U.S.A had the largest amount of competitors
compared to other countries, this could be a reason for the large growth in those years.

# What Role Does Age Play (Section 4)

Strength is a characteristic which is developed over time, meaning it takes time for
an individuals body to reach its maximum potential. Unfortunately, the body also
deteriorates over time. In this section, we explore this relationship with hopes in
finding the age class where the strength is at its highest. To accomplish this, we
will examine both the male and female average compound lifts (squat, bench, and deadlift)
across each age class. We expect the highest lifts to be found in the 24-34 age class
as this is where bodies typically hit their peak with consistent weight training.

# Useful Functions
```{r}
# This function gets the ages of the individuals with the highest totals in each
# age class. It returns them as a data frame.
#
# Args: data - the data frame that will be altered
max_age_class <- function(data){
  index <- data |>
      filter(!is.na(age) & !is.na(age_class)) |>  # remove NA values to save errors
      group_by(age_class) |> # create our groups of age classes
      filter((total_kg == max(total_kg, na.rm = TRUE))) |> # get the max total for each age class
      select(name, age, age_class, total_kg) |> # select the columns we want
      distinct(name, age, age_class, total_kg) |> # remove all duplicate values
      arrange(age_class)
  return (index)
}
```

# Male Strength by Age Class
```{r}
# this is the order of the age classes from youngest to oldest
age_levels = c("5-12", "13-15", "16-17", "18-19", "20-23", "24-34", "35-39", "40-44", 
               "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80-999")

# get only the men, and turn the age_class into a factor for organization
male_averages <- ipf |> filter(sex == "M") |>
                mutate(age_class = factor(age_class, levels = age_levels))

# now we can perform our actual calculations for men
male_result <- avg_weights(male_averages, age_class)
male_result
```

# Graph Male Average Strength
```{r}
# create a scatter plot with lines which illustrates the average lift weight for 
# the various male weight classes 
male_result |> ggplot(aes(x = age_class)) +
  geom_point(aes(y = avg_squat, col = "Squat", shape = "Squat"), size = 2.5) +
  geom_point(aes(y = avg_bench, col = "Bench", shape = "Bench"), size = 2.5) +
  geom_point(aes(y = avg_dead, col = "Deadlift", shape = "Deadlift"), size = 2.5) +
  geom_line(aes(y = avg_squat, color = "Squat", group = 1), linetype = "solid") +
  geom_line(aes(y = avg_bench, color = "Bench", group = 1), linetype = "solid") +
  geom_line(aes(y = avg_dead, color = "Deadlift", group = 1), linetype = "solid") +
  scale_color_manual(values = c("Squat" = "blue", "Bench" = "red", "Deadlift" = "darkgreen")) +
  scale_shape_manual(values = c("Squat" = 1, "Bench" = 2, "Deadlift" = 3)) +
  labs(
    title = "Average Male Squat, Bench, & Deadlift Across Age Classes",
    x = "Age Class",
    y = "Weight (kg)",
    color = "Exercise Type",
    shape = "Exercise Type"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

# Female Strength by Age Class
```{r}
# get only the females, and turn the age_class into a factor for organization
female_averages <- ipf |> filter(sex == "F") |>
                mutate(age_class = factor(age_class, levels = age_levels))

# now we can perform our actual calculations for female
female_result <- avg_weights(female_averages, age_class)
female_result
```

# Graph Female Strength by Age Class
```{r}
# create a scatter plot with lines which illustrates the average lift weight for 
# the various male weight classes 
female_result |> ggplot(aes(x = age_class)) +
  geom_point(aes(y = avg_squat, col = "Squat", shape = "Squat"), size = 2.5) +
  geom_point(aes(y = avg_bench, col = "Bench", shape = "Bench"), size = 2.5) +
  geom_point(aes(y = avg_dead, col = "Deadlift", shape = "Deadlift"), size = 2.5) +
  geom_line(aes(y = avg_squat, color = "Squat", group = 1), linetype = "solid") +
  geom_line(aes(y = avg_bench, color = "Bench", group = 1), linetype = "solid") +
  geom_line(aes(y = avg_dead, color = "Deadlift", group = 1), linetype = "solid") +
  scale_color_manual(values = c("Squat" = "blue", "Bench" = "red", "Deadlift" = "darkgreen")) +
  scale_shape_manual(values = c("Squat" = 1, "Bench" = 2, "Deadlift" = 3)) +
  labs(
    title = "Average Female Squat, Bench, & Deadlift Across Age Classes",
    x = "Age Class",
    y = "Weight (kg)",
    color = "Exercise Type",
    shape = "Exercise Type"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```
# Strongest Age in Each Age Class
```{r}
# First we find the strongest males age in each age class
strongest_males <- max_age_class(male_averages)
strongest_males

# Next we find the strongest females age in each age class
strongest_females <- max_age_class(female_averages)
strongest_females
```

# Conclusion for Section 4
As we examine the relationship between age and strength there are a few observations that can
be made. First of all, our guess for the strongest age class being 24-34 is generally true. For
the female section it is evident that this is the case; however, as we examine the male section
we see that the strongest age class is spread across two age classes. Both the male 24-34 and
35-39 age classes have similar lift numbers, from which we can deduce that the male have a slightly
longer peak than the females do.

Looking into this relationship further, we decided to find the age of the individual with the highest 
total in each age class. This was once again split between male and female. By doing this, we see the
age of the individual with the highest total is typically near the upper limit of the age class for the
age classes prior to 24-34. After this age class, the highest total was generally held by individuals
with an age closer to the lower limit of the weight class. These results may be slightly skewed, as
contestants with an age or age class of NA were not included, so their may be higher totals that were
not accounted for. Overall, there is a clear correlation between age and strength.

# Difference in equipment types. (Section 5)
 


# Conclusion