---
title: "Powerlifting: An Exploration on the Data"
author: "Zach Eanes & Dalton Rogers"
output: html_document
date: "2024-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

ipf <- read_csv("openipf-2024-04-13.csv")
# View(ipf)

ipf <- janitor::clean_names(ipf)
```

# What is "Powerlifting"?
As an easy description, powerlifting is the open sport of competing to see who can lift
the most total weight across three primary lifts. These lifts being Squat, Bench, and Deadlift 
(referred to as SBD throughout this project). 

The general structure of many of these competitions includes many little factors:
1) Everyone is split into weight classes, gender, and age classes. This is to ensure everyone is 
   competing fairly and no one has a clear advantage.
2) Every lifter has 3 attempts for each lifting, leaving 9 total lifts across the competition. 
3) Drug free! No steroids or substances that may boost someone's performance, and these 
   competitors are tested to ensure they are drug free.
4) The concept of good and failed lifts: There's 3 judges for a lift, all viewing a different 
   perspective of the lifter. In order for a lift to be good, 2/3 judges must deem it a good lift.
   Some common failures include not getting enough depth squatting, not controlling the drop after
   a deadlift, or of course literally not getting the weight up.
   
# Linear Progression of Weight Classes?
This one is simple, do heavier weights classes generally lift higher amounts than lower? We can
observe the best successful lift for each SBD and find an average for each weight class group
and then we can compare the classes to see if there's a linear progression.

IPF lists their male weight classes as:
59.0 kg Class:   up to 59.0 kg
66.0 kg Class:   from 59.01 kg up to 66.0 kg
74.0 kg Class:   from 66.01 kg up to 74.0 kg
83.0 kg Class:   from 74.01 kg up to 83.0 kg
93.0 kg Class:   from 83.01 kg up to 93.0 kg
105.0 kg Class:  from 93.01 kg up to 105.0 kg
120.0 kg Class:  from 105.01 kg up to 120.0 kg
120.0+ kg Class: from 120.01 kg up to unlimited

IPF lists their female weifht classes as:
47.0 kg Class:  up to 47.0 kg
52.0 kg Class:  from 47.01 kg up to 52.0 kg
57.0 kg Class:  from 52.01 kg up to 57.0 kg
63.0 kg Class:  from 57.01 kg up to 63.0 kg
69.0 kg Class:  from 63.01 kg up to 69.0 kg
76.0 kg Class:  from 69.01 kg up to 76.0 kg
84.0 kg Class:  from 76.01 kg up to 84.0 kg
84.0+ kg Class: from 84.01 kg up to unlimited
```{r avg_success}
# get those in proper weight classes, and make weight_class_kg a factor for organization
male_factors = c(59, 66, 74, 83, 93, 105, 120, "120+")
male_classes <- ipf |> 
    filter(weight_class_kg %in% c(59, 66, 74, 83, 93, 105, 120, "120+") & sex == "M") |>
    mutate(weight_class_kg = factor(weight_class_kg, levels = male_factors))

female_factors = c(47, 52, 57, 63, 69, 76, 84, "84+")
female_classes <- ipf |>
    filter(weight_class_kg %in% c(47, 52, 57, 63, 69, 76, 84, "84+") & sex == "F") |>
    mutate(weight_class_kg = factor(weight_class_kg, levels = female_factors))
           


# This function gets the average of the best lift for each factor and arranges it
# by the weight class itself
#
# Args: class - the class we want to get the average weights for
avg_weights <- function(class) {
    class |> 
    filter(!is.na(best3squat_kg) & !is.na(best3bench_kg) & !is.na(best3deadlift_kg)) |> 
        group_by(weight_class_kg) |> # group by their respective classes
        summarise(
            avg_squat = sum(best3squat_kg) / n(), # get avg squat
            avg_bench = sum(best3bench_kg) / n(), # get avg bench
            avg_dead = sum(best3deadlift_kg) / n()) |> # get avg deadlift 
        arrange(weight_class_kg) # arrange to make it better
}

# This function takes our data and makes the data itself tidier and simple to use for
# both data exploration and graphing.
#
# Args: class - the class we want to make tidier 
make_avgs_tidyer <- function(class) {
    # pivot to long format to graph easier
    class |> pivot_longer(cols = c(avg_squat, avg_bench, avg_dead),
                          names_to = "lift", values_to = "avg_weight")
}

# get information we need for our data and make it usable
weights_male <- avg_weights(male_classes) |> make_avgs_tidyer(); weights_male
weights_female <- avg_weights(female_classes) |> make_avgs_tidyer(); weights_female 

# group by the weight class and put sbd all next to each other in group
weights_male |> ggplot(aes(x = weight_class_kg, y = avg_weight, fill = lift)) + 
    geom_bar(stat = "identity", position = "dodge") +
    labs(
        title = "Comparing Male Average Weights for SBD across Common Weight Classes",
        x = "Weight Class (kg)",
        y = "Average Weight (kg)"
    ) + scale_fill_discrete(name = "What Lift?", labels = c("Bench", "Deadlift", "Squat"))

weights_female |> ggplot(aes(x = weight_class_kg, y = avg_weight, fill = lift)) + 
    geom_bar(stat = "identity", position = "dodge") +
    labs(
        title = "Comparing Female Average Weights for SBD across Common Weight Classes",
        x = "Weight Class (kg)",
        y = "Average Weight (kg)"
    ) + scale_fill_discrete(name = "What Lift?", labels = c("Bench", "Deadlift", "Squat"))
```
In order to get this information together, we had many failures and attempts at using the 
data is a less tidy manner. The biggest change was the inclusion of the pivot_longer in order
to allow us to create more rows and make the graphing easier. It allows us to group and place
each lift inside of the group, as well as creating a cleaner chart in general. Beforehand,
we broke some of the basic policies of tidy data, including storing too much information
inside of a cell and "double dipping". We had numbers in the columns for the lift itself, and
in the end out newest form is tidiest and surely best for our purposes. Below is some of the 
attempts we had beforehand, which ended as a failure in all honesty.

```{r avg_failure}
# below method is an early attempt that ended in failure 
pre_avg_weights <- function(class) {
    class |>  # filter out any entry missing the best lift
        filter(!is.na(best3squat_kg) & !is.na(best3bench_kg) & !is.na(best3deadlift_kg)) |> 
        group_by(weight_class_kg) |> # group by their respective classes
        summarise(
            avg_squat = sum(best3squat_kg) / n(), # get avg squat
            avg_bench = sum(best3bench_kg) / n(), # get avg bench
            avg_dead = sum(best3deadlift_kg) / n()) |> # get avg deadlift 
        arrange(weight_class_kg)
}

# recreate our data set for demonstration purposes
male_classes <- ipf |> 
    filter(weight_class_kg %in% c(59, 66, 74, 83, 93, 105, 120, "120+") & sex == "M") |>
    mutate(weight_class_kg = factor(weight_class_kg, 
                                       levels = male_factors))
# this was the dataset before
pre_avg_weights(male_classes) 

# this was out plot before the pivot_longer, as we just had each lift in thier own graph
pre_avg_weights(male_classes) |> ggplot(aes(x = weight_class_kg, y = avg_squat, fill = avg_squat)) + 
    geom_bar(stat = "identity") 

```


# Percentage of Failed Lifts Across all Nine Attempts
In this section, we wanted to observe if for certain lifts, competitors are more likely to
fail the lift. In the database, if a life begins with a "-", this means it was a failure. 
Using this, we may be able to get the percentage of good lifts to failed lifts for each lift, 
all 3 for SBD.
```{r}
# first we remove bad entries, where any of their lifts are na. sum up only data we want
# to observe
good_lifts <- ipf |> 
    # filter our bad entries
    filter(!is.na(squat1kg) & !is.na(squat2kg) & !is.na(squat3kg) &
           !is.na(bench1kg) & !is.na(bench2kg) & !is.na(bench3kg) &
           !is.na(deadlift1kg) & !is.na(deadlift2kg) & !is.na(deadlift3kg)) |> 
    # only get what we need
    summarize(name, squat1kg, squat2kg, squat3kg, bench1kg, bench2kg, bench3kg,
              deadlift1kg, deadlift2kg, deadlift3kg)
    
# pivot longer to get the data as column lift with a value of the column name and weight 
# as a value in another
good_lifts <- good_lifts |> 
    # split the dataset into lift and weight, to observe entries as group easier
    pivot_longer(cols = c(squat1kg, squat2kg, squat3kg, bench1kg, bench2kg, bench3kg, 
                          deadlift1kg, deadlift2kg, deadlift3kg),
                 names_to = "lift", values_to = "weight")

# this function gets the percentage of failed lifts for each lift. because the dataset 
# lists the failed lifts beginning with a -, we can use regex to explore failure 
# to success
#
# Args: class - the data we want to get the percentage of failed lifts for
percentage_failed <- function(class) {
    class |> 
        # group by lifts to see the success/fail rate of each
        group_by(lift) |>
        # get the percentage of failed lifts for each lift
        summarise( # get the lift and their respective failure/success rate
            failed = sum(str_detect(weight, "^-.*")) / n() * 100,
            num_failed = sum(str_detect(weight, "^-.*")),
            success = 100 - failed,
            num_success = n() - num_failed,
            total = failed + success
        )
}

# get the percentage of failed lifts for each lift
rates <- percentage_failed(good_lifts)

# longer data in order to graph easier :)
lift_data_long <- rates |> 
    pivot_longer(cols = c(failed, success), names_to = "outcome", values_to = "percentage")

# Plot
ggplot(lift_data_long, aes(x = lift, y = percentage, fill = outcome)) +
    # bar chart of the percentage of failed and successful lifts for each lift
    geom_bar(stat = "identity") + 
    # add text to the chart showing the percentage itself 
    geom_text(aes(label = paste0(round(percentage, 2), "%")), 
              position = position_stack(vjust = 0.5), 
              color = "white") +
    labs(
        title = "Percentage of Failed and Successful Lifts for Each Lift",
        x = "Lift",
        y = "Percentage (%)",
        fill = "Outcome"
    ) +
    # colors for the percentages themselves 
    scale_fill_manual(values = c("success" = "darkgreen", "failed" = "red"), 
                      name = "Outcome of Lift", labels = c("Failure", "Success")) +
    # make the x axis text easier to read
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
```
Looking at the chart itself, there's some shocking results. First, we never expected the last
bench attempt to have the highest failure rate. This is because the bench press is generally
attempted safely, as competitors still have the deadlift after to try and get more. Our guess was
that the last deadlift would have the highest failure rate, for many reasons:
1) It's the last lift, so competitors will be more fatigued from the rest of the event. 
2) Since it's the last, many competitors go "all-out" to attempt to place on the podium.
3) In generality, deadlift is the hardest lift to get up, because grip strength is a large 
   limiting factor for the strongest competitors.
   
Lastly, we were able to conclude that there is some trend to where the first lift is the most
successful, and the last lift is the least successful. This is because the first lift is generally
the lightest, and the last lift is generally the heaviest. This is a trend we expected, but it's
nice to see it confirmed in the data.


